Some examples in extended logic counting classes

### 2015-12-15 tex/examples.tex
2 nov
15 dic

\subsection{{\sc \#3-DNF}}
We will now show that {\sc \#3-DNF} $\in\EoneE$.\\

@@

\subsection{{\sc \#3-Horn-Sat}}
We will now show that {\sc \#3-Horn-Sat} $\in\EoneU$.\\

@@

\subsection{{\sc \#DNF}}
We will now show that {\sc \#DNF} $\in\EtwoE$.\\

@@

By extension, {\sc \#3-DNF} $\in\EtwoE$.

\subsection{{\sc \#Horn-Sat}}
We will now show that {\sc \#Horn-Sat} $\in\EoneU$.\\

@@

\subsection{{\sc \#Disj-Horn-Sat}}
We will now show that {\sc \#Disj-Horn-Sat} $\in\EtwoE$.

>>>>

\subsection{{\sc \#3-DNF}}
We will now show that {\sc \#3-DNF} $\in\XE{1}$.\\

@@

\subsection{{\sc \#3-Horn-Sat}}
We will now show that {\sc \#3-Horn-Sat} $\in\XU{1}$.\\

@@

\subsection{{\sc \#DNF}} \label{dnfdef}
We will now show that {\sc \#DNF} $\in\XE{2}$.\\

@@

By extension, {\sc \#3-DNF} $\in\XE{2}$.

\subsection{{\sc \#Horn-Sat}}
We will now show that {\sc \#Horn-Sat} $\in\XU{1}$.\\

@@

\subsection{{\sc \#Disj-Horn-Sat}}
We will now show that {\sc \#Disj-Horn-Sat} $\in\XE{2}$.

### 2015-12-15 tex/first_h.tex
15 dic

We define syntactically the classes $\Eie$ and $\Eiu$ according to the following grammar:
\begin{enumerate}
	\item $\Eezero$:
	\[
		E_0 ::= \varphi , \varphi \mbox{ is an {\sc FO}-extended quantifier-free $\L$-formula }.
	\]
	\item $\Euzero$:
		\[U_0 ::= E_0.\]
	\item $\Eipe$:
	\[	E_{i+1} ::= U_i \ \mid \ \exists x \, E_{i+1}.\]
	\item $\Eipu$:
		\[U_{i+1} ::= E_i \ \mid \ \forall x \, U_{i+1}.\]
\end{enumerate}

A function $f$ is in $\Eie$ (resp. $\Eiu$) if there is an $\L$-formula $\varphi$ defined by the grammar $E_i$ (resp. $U_i$) such that $f = f_{\varphi}$.

\begin{theo} \label{fp1}
	$\Eo \subseteq$ {\sc FP}.
\end{theo}
\begin{proof}
	Let $f \in \Eo$, and let $\varphi(\X,\x)$ be and {\sc FO}-extended quantifier-free $\L$-formula such that:
	\begin{eqnarray*}
		f(\A) &=& \vert \{\langle \P,\e  \rangle \mid \A \models \varphi(\P,\e) \} \vert
	\end{eqnarray*}
	for each $\A = \langle A, \S^{\A}, \leq^{\A} \rangle \in \Truc$, where $\e \in A^m$ and $\P = (P_1,\ldots,P_q)$ is a tuple of predicates. We will now show that computing $f(\A)$ can be done in polynomial time.

@@

		($\Leftarrow$) Suppose $g(\A) > 0$. Let $\Q$ and $\c = (\c_1,\c_2)$, where $\c_1$ and $\c_2$ have $k$ and $m$ elements respectively, be such that $\A \models \psi(\Q,\c) = \varphi(\c_1,\Q,\c_2)$. Then we have that $\A \models \exists \x \, \varphi(\x,\Q,\c_2)$, from which we conclude that $f(\A) > 0$.
	\end{proof}
	Note that $\psi(\X,\y)$ is an {\sc FO}-extended $\L$-formula, so $g \in\Eo$. It follows from \ref{mono} that for each $\A \in \Truc$, it holds that $\A \in L_f$ if and only if $g(\A) > 0$. But by Theorem \ref{fp1}, $g(\A)$ can be evaluated in polynomial time. From this we conclude that the decision version of $f$ is in $\ptime$.
\end{proof}

@@

		Let $i$ be such that $\A\models \exists\x\,\varphi_i(\x,\P,\e)$. It is simple to see that from the way we constructed $\Phi_i^{\e}$, it holds that $\sigma(\Phi_i^{\e}) = 1$. Also, note that every different assignment $\langle \P',\e' \rangle$, it holds that $\sigma(\langle \P',\e' \rangle) \neq \sigma(\langle \P,\e \rangle)$. Therefore, $g$ is an injective function. Now, let $\sigma$ be an arbitrary assignment such that $\sigma(\Phi) = 1$. Let $i$ and $\e$ be such that $\sigma(\Phi^{\e}_i)$. Note that for every $\e'\in A^d\setminus\{\e\}$, it holds that $\sigma(\theta_{\e'}) = 0$, so there is exactly one possible $\e$ associated to $\sigma$. Consider the assignment $\P$ where $\b\in P_{\ell}$ if and only if $\sigma(p^{\e}_{\ell,\b}) = 1$, for each $\ell$ and $\b \in A^{c_{\ell}}$. Since there is at least one $\a\in\Gamma_i$ for which $\sigma$ satisfies its respective subformula, it holds that $\A\models\varphi_i(\a,\P,\e)$, which implies that $\A\models\exists\x\,\varphi(\x,\P,\e)$. From this we deduce that $g$ is surjective, and therefore a bijection. We conclude that $f(\A) = f_{\textsc{\#DNF}}(\A_{\Phi})$.
	\end{enumerate}
\end{proof}
\begin{theo}
	$\SUo$ with $n$ open first-order variables is properly contained in $\SUo$ with $n+1$ open first-order variables for $n\in\N$.  
\end{theo}
\begin{proof}
	Suppose it is not properly contained for $n$. Let $\z = (z_1,\ldots,z_{n+1})$, let $\u = (u_1,\ldots,u_n)$ and let $\varphi(\z) = \bigwedge_{i = 1}^{n} (z_i < z_{i+1} \wedge \forall y(z_i\not<y \vee y\not<z_{i+1} )) \wedge \forall y(y \not< z_1)$. Note that $f_{\varphi(\z)}(\A) = 1$ if $\A$ has at least $n+1$ elements, and  $f_{\varphi(\z)}(\A) = 0$ otherwise, for each $\A\in\Truc$. Let $\forall\x\,\psi(\x,\X,\u)$ be such that $f_{\varphi(\z)} = f_{\forall\x\,\psi(\x,\X,\u)}$. Let $\A = \langle A, \S^{\A}, <^{\A} \rangle\in\Truc$ where $\vert A \vert = n+1$. Note that $f_{\forall\x\,\psi(\x,\X,\u)}(\A) = 1$. Let $(\P,\e)$ be such that $\A\models\forall\x\,\psi(\x,\P,\e)$. Let $a\in A$ be such that $a\not\in\e$. Consider $\B$ as the induced substructure of $\A$ with domain $A\setminus\{a\}$, and let $\Q$ be the corresponding induced sub-assignment of $\P$. As universal formulae are preserved under substructure, we have that $\A\models\forall\x\,\psi(\x,\Q,\e)$. This implies that $f_{\forall\x\,\psi(\x,\X,\u)}(\B) \geq 1$, which leads to a contradiction.
\end{proof}

>>>>

We define syntactically the classes $\XE{i}$ and $\XU{i}$ according to the following grammar:
\begin{enumerate}
	\item $\XE{0}$:
	\[
		E_0 ::= \varphi , \varphi \mbox{ is an {\sc FO}-extended quantifier-free $\L$-formula }.
	\]
	\item $\XU{0}$:
		\[U_0 ::= E_0.\]
	\item $\XE{i+1}$:
	\[	E_{i+1} ::= U_i \ \mid \ \exists x \, E_{i+1}.\]
	\item $\XU{i+1}$:
		\[U_{i+1} ::= E_i \ \mid \ \forall x \, U_{i+1}.\]
\end{enumerate}

A function $f$ is in $\XE{i}$ (resp. $\XU{i}$) if there is an $\L$-formula $\varphi$ defined by the grammar $E_i$ (resp. $U_i$) such that $f = f_{\varphi}$.

\begin{theo} \label{fp1}
	$\XE{0} \subseteq$ {\sc FP}.
\end{theo}
\begin{proof}
	Let $f \in \XE{0}$, and let $\varphi(\X,\x)$ be and {\sc FO}-extended quantifier-free $\L$-formula such that:
	\begin{eqnarray*}
		f(\A) &=& \vert \{\langle \P,\e  \rangle \mid \A \models \varphi(\P,\e) \} \vert
	\end{eqnarray*}
	for each $\A = \langle A, \S^{\A}, \leq^{\A} \rangle \in \Truc$, where $\e \in A^m$ and $\P = (P_1,\ldots,P_q)$ is a tuple of predicates. We will now show that computing $f(\A)$ can be done in polynomial time.

@@

		($\Leftarrow$) Suppose $g(\A) > 0$. Let $\Q$ and $\c = (\c_1,\c_2)$, where $\c_1$ and $\c_2$ have $k$ and $m$ elements respectively, be such that $\A \models \psi(\Q,\c) = \varphi(\c_1,\Q,\c_2)$. Then we have that $\A \models \exists \x \, \varphi(\x,\Q,\c_2)$, from which we conclude that $f(\A) > 0$.
	\end{proof}
	Note that $\psi(\X,\y)$ is an {\sc FO}-extended $\L$-formula, so $g \in\XE{0}$. It follows from \ref{mono} that for each $\A \in \Truc$, it holds that $\A \in L_f$ if and only if $g(\A) > 0$. But by Theorem \ref{fp1}, $g(\A)$ can be evaluated in polynomial time. From this we conclude that the decision version of $f$ is in $\ptime$.
\end{proof}

@@

		Let $i$ be such that $\A\models \exists\x\,\varphi_i(\x,\P,\e)$. It is simple to see that from the way we constructed $\Phi_i^{\e}$, it holds that $\sigma(\Phi_i^{\e}) = 1$. Also, note that every different assignment $\langle \P',\e' \rangle$, it holds that $\sigma(\langle \P',\e' \rangle) \neq \sigma(\langle \P,\e \rangle)$. Therefore, $g$ is an injective function. Now, let $\sigma$ be an arbitrary assignment such that $\sigma(\Phi) = 1$. Let $i$ and $\e$ be such that $\sigma(\Phi^{\e}_i)$. Note that for every $\e'\in A^d\setminus\{\e\}$, it holds that $\sigma(\theta_{\e'}) = 0$, so there is exactly one possible $\e$ associated to $\sigma$. Consider the assignment $\P$ where $\b\in P_{\ell}$ if and only if $\sigma(p^{\e}_{\ell,\b}) = 1$, for each $\ell$ and $\b \in A^{c_{\ell}}$. Since there is at least one $\a\in\Gamma_i$ for which $\sigma$ satisfies its respective subformula, it holds that $\A\models\varphi_i(\a,\P,\e)$, which implies that $\A\models\exists\x\,\varphi(\x,\P,\e)$. From this we deduce that $g$ is surjective, and therefore a bijection. We conclude that $f(\A) = f_{\textsc{\#DNF}}(\A_{\Phi})$.
	\end{enumerate}
\end{proof}
\begin{theo}
	$\U{1}$ with $n$ open first-order variables is properly contained in $\U{1}$ with $n+1$ open first-order variables for $n\in\N$.  
\end{theo}
\begin{proof}
	Suppose it is not properly contained for $n$. Let $\z = (z_1,\ldots,z_{n+1})$, let $\u = (u_1,\ldots,u_n)$ and let $\varphi(\z) = \bigwedge_{i = 1}^{n} (z_i < z_{i+1} \wedge \forall y(z_i\not<y \vee y\not<z_{i+1} )) \wedge \forall y(y \not< z_1)$. Note that $f_{\varphi(\z)}(\A) = 1$ if $\A$ has at least $n+1$ elements, and  $f_{\varphi(\z)}(\A) = 0$ otherwise, for each $\A\in\Truc$. Let $\forall\x\,\psi(\x,\X,\u)$ be such that $f_{\varphi(\z)} = f_{\forall\x\,\psi(\x,\X,\u)}$. Let $\A = \langle A, \S^{\A}, <^{\A} \rangle\in\Truc$ where $\vert A \vert = n+1$. Note that $f_{\forall\x\,\psi(\x,\X,\u)}(\A) = 1$. Let $(\P,\e)$ be such that $\A\models\forall\x\,\psi(\x,\P,\e)$. Let $a\in A$ be such that $a\not\in\e$. Consider the vocabulary $\L' = \L \cup \{\Q\}$, and let $\A' = \A \cup \langle \Q^{\A'} \rangle$ where $\Q^{\A'} = \P$. Note that $\A\models\varphi(\P,\e)$ if and only if $\A'\models\varphi(\e)$. Consider $\B'$ as the induced substructure of $\A'$ with domain $A\setminus\{a\}$, and let $\Q^{\B'}$ be the induced interpretation of $\Q$. Also let $\B$ be the induced substructure of $\A$ with domain $A\setminus\{a\}$. As universal formulae are preserved under substructure\footnote{Hodges, 1997. P. 183}, we have that $\B'\models\forall\x\,\psi(\x,\e)$, which implies that $\B\models\forall\x\,\psi(\x,\Q^{\B'},\e)$. We conclude that $f_{\forall\x\,\psi(\x,\X,\u)}(\B) \geq 1$, which leads to a contradiction.
\end{proof}

### 2015-12-15 tex/horn_h.tex
15 dic

\section{The Extended Horn Counting Hierarchy}

We define syntactically the classes $\EiE$ and $\EiU$ as follows. First we define extended Horn clauses.
\begin{eqnarray*}
PL &::=& X_i(\x),\, i\in\N,\\
NL &::=& \neg X_i(\x),\, i\in\N \ \mid \ \exists x\, NL,\\
NC &::=& NL \ \mid \ \alpha, \alpha \mbox{ is an {\sc FO}-formula over } \L \ \mid \ (NC \vee NC),\\
HC &::=& NC \ \mid \ (NC \vee PL),
\end{eqnarray*}
where $\x$ is a tuple with the corresponding number of variables. Now we define the syntax of the classes inductively.
\begin{enumerate}
	\item $\EzeroE$:
	\begin{eqnarray*}
	E_0 &::=& HC \ \mid \ E_0 \wedge E_0.
	\end{eqnarray*}
	\item $\EzeroU$:
	\begin{eqnarray*}
	U_0 &::=& E_0.
	\end{eqnarray*}
	\item $\EipE$:
	\begin{eqnarray*}
	E_{i+1} &::=& U_i \ \mid \ \exists x \, E_{i+1}.
	\end{eqnarray*}
	\item $\EipU$:
	\begin{eqnarray*}
		U_{i+1} &::=& E_i \ \mid \ \forall x \, U_{i+1}.
	\end{eqnarray*}
\end{enumerate}
A function $f$ is in $\EiE$ (resp. $\EiU$) if there is an $\L$-formula $\varphi$ defined by the grammar $E_i$ (resp. $U_i$) such that $f = f_{\varphi}$.

The class $\PE$ is defined in [?] as $\PE = \{f\mid f\in\#P \mbox{ and its decision version } L_{f}\in P\}$.

\begin{theo}
	$\EtwoE \subseteq \PE$
\end{theo} 
\begin{proof}
	Let $f = f_{\varphi(\X,\z)}$ such that $\varphi(\X,\z)$ is defined by the grammar $E_2$. First we notice that, as stated in \cite{DBLP:journals/jcss/SalujaST95}, for every $\L$-formula $\varphi$, $f_{\varphi}\in\#P$.\\

@@

	Now we notice that $\zeta = \exists\X\exists\x\exists\y\,\theta(\X,\x,\y)$ is an existential second-order FO-formula, for which $\A\models\zeta$ can be decided in polynomial time.
\end{proof}

@@

\begin{theo}
	$\textsc{Disj-Horn-Sat}$ is hard for $\EtwoE$ under parsimonious reductions. 
\end{theo} 
\begin{proof}
	Let $f$ be an arbitrary function in $\EtwoE$ and let $f_{\textsc{Disj-Horn-Sat}}$ be the function that characterizes {\sc \#Disj-Horn-Sat}. We will now show a function $h:\Truc\to L(P)$ such that for every $\A\in\Truc$, it holds that $f(\A) = f_{\textsc{Disj-Horn-Sat}}(h(\A))$, and which can be computed in polynomial time. Let $\psi$ be such that $f = f_{\psi}$. We separate the proof in three cases:\begin{enumerate}
		\item All free variables in $\psi$ are of first order. Let $\psi = \exists \x \forall \y\,\varphi(\x,\y,\z)$ where $\z = (z_1,\ldots,z_d)$. Given a structure $\A = \langle A, \S^{\A}, \leq^{\A} \rangle\in\Truc$, we generate a propositional formula $\Phi = h(\A)$. First we notice that this case is trivial since we can calculate this function in polynomial time, by checking for each $\e\in A^d$ if $\A\models\exists\x\forall \y\,\varphi(\x,\y,\e)$. Let $n$ be the number of such assignments to $\z$. Let $\p$ be an $n$-tuple of propositional variables where $\p = (p_1,\ldots,p_n)$. The Disj-Horn-Sat formula we return is
		\[
		\Phi = \bigvee_{i = 1}^n \neg p_1 \wedge \cdots \wedge \neg p_{i-1} \wedge p_i \wedge \neg p_{i+1} \wedge \cdots \wedge \neg p_n
		\]
		which has exactly $n$ possible assignments. Note that each $p_i$ and $\neg p_i$ is a Horn clause itself.

>>>>

\section{The Horn Counting Hierarchy}

We define the syntactic classes $\XE{i}$ and $\XU{i}$ as follows. First we define extended Horn clauses.
\begin{eqnarray*}
PL &::=& X_i(\x),\, i\in\N,\\
NL &::=& \neg X_i(\x),\, i\in\N \ \mid \ \exists x\, NL,\\
NC &::=& NL \ \mid \ \alpha, \alpha \mbox{ is an {\sc FO}-formula over } \L \ \mid \ (NC \vee NC),\\
HC &::=& NC \ \mid \ (NC \vee PL) \ \mid \ PL,
\end{eqnarray*}
where $PL$ represents a positive literal, $NL$ is an \textit{extended} negative literal, $NC$ is an extended Horn clause, and $HC$ is an extended Horn formula. Now we define the syntax of the classes inductively.\textbf{}
\begin{enumerate}
	\item $\HE{0}$:
	\begin{eqnarray*}
	E_0 &::=& HC \ \mid \ E_0 \wedge E_0.
	\end{eqnarray*}
	\item $\HU{0}$:
	\begin{eqnarray*}
	U_0 &::=& E_0.
	\end{eqnarray*}
	\item $\HE{i+1}$:
	\begin{eqnarray*}
	E_{i+1} &::=& U_i \ \mid \ \exists x \, E_{i+1}.
	\end{eqnarray*}
	\item $\HU{i+1}$:
	\begin{eqnarray*}
		U_{i+1} &::=& E_i \ \mid \ \forall x \, U_{i+1}.
	\end{eqnarray*}
\end{enumerate}
A function $f$ is in $\XE{i}$ (resp. $\XU{i}$) if there is an $\L$-formula $\varphi$ defined by the grammar $E_i$ (resp. $U_i$) such that $f = f_{\varphi}$.

The class $\PE$ is defined in \cite{DBLP:conf/mfcs/PagourtzisZ06} as $\PE = \{f\mid f\in\#P \mbox{ and its decision version } L_{f}\in P\}$.

\begin{theo}
	$\HE{2} \subseteq \PE$
\end{theo} 
\begin{proof}
	Let $f = f_{\varphi(\X,\z)}$ such that $\varphi(\X,\z)$ is defined by the grammar $E_2$. First we notice that, as stated in \cite{DBLP:journals/jcss/SalujaST95}, for every $\L$-formula $\varphi$, $f_{\varphi}\in\#P$.\\

@@

	Now we notice that $\zeta = \exists\X\exists\x\exists\y\,\theta(\X,\x,\y)$ is an existential second-order FO-formula, for which $\A\models\zeta$ can be decided in polynomial time \cite{DBLP:conf/stacs/Gradel91}.
\end{proof}

@@

\begin{theo}
	$\textsc{Disj-Horn-Sat}$ is hard for $\XE{2}$ under parsimonious reductions. 
\end{theo} 
\begin{proof}
	Let $f$ be an arbitrary function in $\XE{2}$ and let $f_{\textsc{Disj-Horn-Sat}}$ be the function that characterizes {\sc \#Disj-Horn-Sat}. We will now show a function $h:\Truc\to L(P)$ such that for every $\A\in\Truc$, it holds that $f(\A) = f_{\textsc{Disj-Horn-Sat}}(h(\A))$, and which can be computed in polynomial time. Let $\psi$ be such that $f = f_{\psi}$. We separate the proof in three cases:\begin{enumerate}
		\item All free variables in $\psi$ are of first order. Let $\psi = \exists \x \forall \y\,\varphi(\x,\y,\z)$ where $\z = (z_1,\ldots,z_d)$. Given a structure $\A = \langle A, \S^{\A}, \leq^{\A} \rangle\in\Truc$, we generate a propositional formula $\Phi = h(\A)$. First we notice that this case is trivial since we can calculate this function in polynomial time, by checking for each $\e\in A^d$ if $\A\models\exists\x\forall \y\,\varphi(\x,\y,\e)$. Let $n$ be the number of such assignments to $\z$. Let $\p$ be an $n$-tuple of propositional variables where $\p = (p_1,\ldots,p_n)$. The Disj-Horn-Sat formula we return is
		\[
		\Phi = \bigvee_{i = 1}^n \neg p_1 \wedge \cdots \wedge \neg p_{i-1} \wedge p_i \wedge \neg p_{i+1} \wedge \cdots \wedge \neg p_n
		\]
		which has exactly $n$ possible assignments. Note that each $p_i$ and $\neg p_i$ is a Horn clause itself.
